Resonant Scattering in a Uniform Sphere with Large Optical Depth

### 2021-12-13 ms.tex
changing \cite{} to \citet{} according to manuscript guidelines

\subsection{Comparison to Monte Carlo}
The Monte Carlo method is used to solve the transfer equation numerically in order to compare the analytic approximation to an ``exact'' solution. This method is valid at all $\tau_0$, being restricted only by its computational demand which grows proportionally to the number of photons used and $\tau_0$. For each simulation, a total of ${\sim}10^6$ photon packets are initialized at a monochromatic source frequency $x_s$ and are allowed to propagate through the sphere until escaping, at which point their positions, outgoing angles, and escape frequencies are tabulated to obtain the spectrum at the surface of the spherical simulation domain. A constant temperature of $T=10^4\ \rm K$ is set for the gas. Frequency redistribution is calculated at each scattering, including the effects of recoil. In the comparisons shown in this section, the raw photon data is binned in frequency to obtain spectra. Further details of the Monte Carlo implementation are discussed in \cite{2017ApJ...851..150H}.

@@

Figure \ref{fig:xinit} shows how the correction $P_{\rm bc}(x)$ (Eq. \ref{eq:prob_spectrum} with $H \rightarrow H_{\rm bc}$) scales with both $\tau_0$ and $x_s$. In this figure, $\sigma_s$ is shifted by integer multiples of $\tau_0$ (Eq. \ref{eq:change_of_variables}) such that the source falls near the peak of the spectrum for each $\tau_0$. For clarity, only the $x > 0$ side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, it is expected that the fractional size of $H_{\rm bc}$ relative to $H_0$ should become smaller with larger optical depths following $(a\tau_0)^{-1/3}$. This factor has been scaled out of the figure such that solutions for different $\tau_0$ and the same $\sigma_s$ should show close agreement in scale on the figure's vertical axis if the relation holds. Indeed, the scaled solutions converge as $\tau_0$ becomes larger, indicating agreement with the $(a\tau_0)^{-1/3}$ scaling. The remaining discrepancy present in the vertical axis for fixed $\sigma_s$ results from $x_{\rm peak}$ becoming close to $x_{\rm cw}$; at lower $\tau_0$, this causes the line profile approximation in the wing (Eq. \ref{eq:app:line_profile_wing}) to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others are indeed proportional to $(a\tau_0)^{-1/3}$.

@@

\subsection{Steady State Source}
A primary goal of this work is to present a solution for resonant scattering of photons near the line-center frequency $\nu_0$ in a uniform sphere. We have generalized a spherically symmetric solution derived by \cite{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center. We introduce a new term to this solution, $J_{\rm bc}$, which allows the boundary condition $J=\sqrt{3}H$ to be satisfied at the surface of the sphere. This is solved using a continuous Fourier expansion in frequency. The integrals are discretized and the Fourier coefficients solved for numerically. The resulting flux correction, $H_{\rm bc}$, scales as $H_0(a\tau_0)^{-1/3}$. Thus, for large $a\tau_0$, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower $a\tau_0$. Since the Laplacian form of the differential equation is only correct when the line profile is $\phi \approx a/(\pi x^2 \Delta)$, our solutions do not accurately model the Doppler core of the \lya line. Because the peak of the spectral energy distribution of escaping photons is $x_{\rm peak} {\sim} (a\tau_0)^{1/3}$, calculations performed at small $a\tau_0$ are inaccurate due to the close proximity of the spectral peak and the Doppler core of the line.

By comparison with Monte Carlo simulations, we have shown that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions provided a sufficiently large $a\tau_0$. Specifically, this solution shows improvement over previous solutions that utilized a $J=0$ surface boundary condition presented in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, and \cite{2006ApJ...649...14D}. Several papers have previously compared these analytic models to Monte Carlo and seen discrepancies on the order of this correction. For example, in the top-left panel of Figure 1 from \cite{2006ApJ...649...14D}, the \lya spectrum emergent from a sphere of uniform optical depth is shown for $\tau_0=10^5, 10^6,$ and $10^7$ at a temperature of $T=10$ K, corresponding to $a=1.5 \times 10^{-2}$. The dotted line showing their theoretically-derived spectrum ($H_0$) displays an excess at the peak of at least 5-10 percent as compared with the Monte Carlo for $\tau_0=10^5$ and $10^6$. Another example is in \citet{2015MNRAS.449.4336S}, where the peak excess in the \lya spectrum is particularly noticeable for line center optical depths of $\tau_0=10^6$ and $10^7$ in the top panel of their Figure 5, which used slab geometry and a gas temperature of $T=10^4$ K. Again, the error in their solution is of order 5-10 percent. Both of these solutions are too large at the spectral peaks and too small further out in the wing, and the error scales approximately as $(a\tau_0)^{-1/3}$. We show in our Figure \ref{fig:sol_mc_tau} that the error present in $H_0$ is corrected by our treatment of the boundary condition at $\tau_0 = 10^7$ for T$=10^4$ K, corresponding to $a=4.72 \times 10^{-4}$. We note that our corrective term $H_{\rm bc}$ is positive in the line wing and negative at the peak of the spectrum, which matches with the discrepancies noted in the aforementioned solutions.

\subsection{Impulsive Source}
The time-dependence of the problem is solved for a $J=0$ boundary condition to characterize the distribution of photon escape times. Solving the differential equation requires a semi-analytic approach, utilizing an expansion in both space and photon frequency. By matching boundary conditions on either side of the photon source frequency $x_s$, the flux at the surface of the sphere is found as a function of $t$ and $\nu$, expressed as a sum over spatial and frequency modes $n$ and $m$, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to broadly agree with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum and emission at the line-center frequency $\nu_0$. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the effect of the Doppler core, which can trap photons at high optical depths until they diffuse outward in frequency, weighting the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. However, an approximate fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution within these constraints.

This leads in to a possible application of this work. Models of the escape of hydrogen gas in exoplanet atmospheres can be constructed with a treatment of resonant scattering of \lya in spherical geometry. The Monte Carlo method can be used for this problem, but is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. We seek to develop a computationally efficient random walk model which requires a solution in spherical geometry. There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations, including core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A} and hybrid diffusion methods \citep{2018MNRAS.479.2065S}, but these methods are not sufficient for our application since they do not track the amount of time photons spend in the line core. Another approach with wide application is modified random walk methods, such as those discussed in \citet{1984JCoPh..54..508F, 2009A&A...497..155M, 2010A&A...520A..70R}; an outgoing photon is randomly sampled on the surface of the outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions, and escape times, based on solutions to the diffusion equation. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \cite{1990ApJ...350..216N} solution, but this solution of course does not use the correct frequency-dependent boundary condition at the surface of the sphere. Furthermore, to perform a full radiation hydrodynamic simulation with Monte Carlo acceleration, it will be necessary to calculate radiation forces due to \lya transfer on the faces of each grid cell in the simulation. Similar calculations have been done in \citet{1976ApJ...208..286W} in plane-parallel geometry. However, these solutions are limited to optical depths below $2.5 \times 10^3$. For this work, it would be necessary to model line center optical depths of up to 1 million or more. We will utilize the solution derived here as the basis for our own novel implementation of the modified random walk method.

@@

We have examined previous solutions to \lya transfer including resonant scattering in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \cite{1973MNRAS.162...43H}, \cite{1990ApJ...350..216N}, \cite{2006ApJ...649...14D} and others does not produce a valid solution to the transfer equation. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition, examining the spectral and time dependence of the resulting equations. The key result is that the errors in the previously-cited works have been quantified via a correction term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}$ is of order unity when the spectral peaks are near the Doppler core, and diminishes at larger $\tau_0$ following a $(a\tau_0)^{-1/3}$ scaling.

Introducing time-dependence to the transfer equation leads to another solution which is solved numerically with an eigenfunction expansion. We demonstrate that it agrees with the steady-state spectrum for $x_s=0$ when integrated over time, though its rate of numerical convergence is slow and requires a sum over many modes to become accurate. This treatment is incomplete for emission away from line center ($x_s\neq0$), however, since any terms that would allow asymmetry in the spectrum are not present in the sum. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of $\tau_0$, noting general agreement in the resulting escape time distributions. In future work, this will be used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.

@@

This research was funded by NASA ATP grant 80NSSC18K0696, ``Exoplanetary MHD Outflows Driven by EUV Heating, Lyman alpha Radiation Forces and Stellar Tides". % TODO: Add Pleiades computer.
\restartappendixnumbering

%TODO: Add code acknowledgements.

>>>>

\subsection{Comparison to Monte Carlo}
The Monte Carlo method is used to solve the transfer equation numerically in order to compare the analytic approximation to an ``exact'' solution. This method is valid at all $\tau_0$, being restricted only by its computational demand which grows proportionally to the number of photons used and $\tau_0$. For each simulation, a total of ${\sim}10^6$ photon packets are initialized at a monochromatic source frequency $x_s$ and are allowed to propagate through the sphere until escaping, at which point their positions, outgoing angles, and escape frequencies are tabulated to obtain the spectrum at the surface of the spherical simulation domain. A constant temperature of $T=10^4\ \rm K$ is set for the gas. Frequency redistribution is calculated at each scattering, including the effects of recoil. In the comparisons shown in this section, the raw photon data is binned in frequency to obtain spectra. Further details of the Monte Carlo implementation are discussed in \citet{2017ApJ...851..150H}.

@@

Figure \ref{fig:xinit} shows how the correction $P_{\rm bc}(x)$ (Eq. \ref{eq:prob_spectrum} with $H \rightarrow H_{\rm bc}$) scales with both $\tau_0$ and $x_s$. In this figure, $\sigma_s$ is shifted by integer multiples of $\tau_0$ (Eq. \ref{eq:change_of_variables}) such that the source falls near the peak of the spectrum for each $\tau_0$. For clarity, only the $x > 0$ side of the spectrum is shown. From Eq. \ref{eq:hbc_scaling}, it is expected that the fractional size of $H_{\rm bc}$ relative to $H_0$ should become smaller with larger optical depths following $(a\tau_0)^{-1/3}$. This factor has been scaled out of the figure such that solutions for different $\tau_0$ and the same $\sigma_s$ should show close agreement in scale on the figure's vertical axis if the relation holds. Indeed, the scaled solutions converge as $\tau_0$ becomes larger, indicating agreement with the $(a\tau_0)^{-1/3}$ scaling. The remaining discrepancy present in the vertical axis for fixed $\sigma_s$ results from $x_{\rm peak}$ becoming close to $x_{\rm cw}$; at lower $\tau_0$, this causes the line profile approximation in the wing (Eq. \ref{eq:app:line_profile_wing}) to break down. From this, we conclude that the errors introduced by the incorrect separation of variables in \citet{1973MNRAS.162...43H}, \citet{1990ApJ...350..216N}, \citet{2006ApJ...649...14D} and others are indeed proportional to $(a\tau_0)^{-1/3}$.

@@

\subsection{Steady State Source}
A primary goal of this work is to present a solution for resonant scattering of photons near the line-center frequency $\nu_0$ in a uniform sphere. We have generalized a spherically symmetric solution derived by \citet{2006ApJ...649...14D} (called $H_0$ here) to allow a monochromatic source of photons with frequencies away from line center. We introduce a new term to this solution, $J_{\rm bc}$, which allows the boundary condition $J=\sqrt{3}H$ to be satisfied at the surface of the sphere. This is solved using a continuous Fourier expansion in frequency. The integrals are discretized and the Fourier coefficients solved for numerically. The resulting flux correction, $H_{\rm bc}$, scales as $H_0(a\tau_0)^{-1/3}$. Thus, for large $a\tau_0$, only a small correction to $H_0$ is needed, while larger errors are present in calculations performed at lower $a\tau_0$. Since the Laplacian form of the differential equation is only correct when the line profile is $\phi \approx a/(\pi x^2 \Delta)$, our solutions do not accurately model the Doppler core of the \lya line. Because the peak of the spectral energy distribution of escaping photons is $x_{\rm peak} {\sim} (a\tau_0)^{1/3}$, calculations performed at small $a\tau_0$ are inaccurate due to the close proximity of the spectral peak and the Doppler core of the line.

By comparison with Monte Carlo simulations, we have shown that the enforcement of the correct frequency-dependent boundary conditions improves the accuracy of these analytic solutions provided a sufficiently large $a\tau_0$. Specifically, this solution shows improvement over previous solutions that utilized a $J=0$ surface boundary condition presented in \citet{1973MNRAS.162...43H}, \citet{1990ApJ...350..216N}, and \citet{2006ApJ...649...14D}. Several papers have previously compared these analytic models to Monte Carlo and seen discrepancies on the order of this correction. For example, in the top-left panel of Figure 1 from \citet{2006ApJ...649...14D}, the \lya spectrum emergent from a sphere of uniform optical depth is shown for $\tau_0=10^5, 10^6,$ and $10^7$ at a temperature of $T=10$ K, corresponding to $a=1.5 \times 10^{-2}$. The dotted line showing their theoretically-derived spectrum ($H_0$) displays an excess at the peak of at least 5-10 percent as compared with the Monte Carlo for $\tau_0=10^5$ and $10^6$. Another example is in \citet{2015MNRAS.449.4336S}, where the peak excess in the \lya spectrum is particularly noticeable for line center optical depths of $\tau_0=10^6$ and $10^7$ in the top panel of their Figure 5, which used slab geometry and a gas temperature of $T=10^4$ K. Again, the error in their solution is of order 5-10 percent. Both of these solutions are too large at the spectral peaks and too small further out in the wing, and the error scales approximately as $(a\tau_0)^{-1/3}$. We show in our Figure \ref{fig:sol_mc_tau} that the error present in $H_0$ is corrected by our treatment of the boundary condition at $\tau_0 = 10^7$ for T$=10^4$ K, corresponding to $a=4.72 \times 10^{-4}$. We note that our corrective term $H_{\rm bc}$ is positive in the line wing and negative at the peak of the spectrum, which matches with the discrepancies noted in the aforementioned solutions.

\subsection{Impulsive Source}
The time-dependent transfer equation is solved in order to characterize the distribution of photon escape times. A semi-analytic approach is used, utilizing an expansion in both space and photon frequency. This boundary value problem is solved to find the flux at the surface of the sphere as a function of $t$ and $\nu$, enforcing matching conditions on either side of the photon source frequency $x_s$. This solution is expressed as a sum over spatial and frequency modes $n$ and $m$, respectively. Calculating additional spatial eigenmodes provides a more accurate model of the physics nearer to line center, but convergence is slow due to each eigenmode's weak dependence on $n$. Additional frequency eigenmodes introduce fast-oscillating terms that improve the accuracy of the sum, as their contributions cancel with components of lower-order terms to better represent the true solution. Integrating the solution over time produces a spectrum that is shown to broadly agree with the steady-state calculations in Section \ref{sec:steadystate}, provided a sufficient number of terms in the sum and emission at the line-center frequency $\nu_0$. Integrating the solution over frequency leads to a distribution of photon escape times, which can be compared directly with Monte Carlo simulations. The sum over eigenmodes produces an escape-time distribution that broadly captures the behavior shown by Monte Carlo data---a rise at early times, transitioning to exponential decay in the tail of the distribution. It is expected that the accuracy of the rate constant for the tail of the distribution is limited by the effect of the Doppler core, which can trap photons at high optical depths until they diffuse outward in frequency, weighting the distribution toward later times. This physics is not modeled by our solution for two reasons: 1) our calculations ignored the Gaussian component of the Voigt line profile, leaving the Lorentzian piece which is accurate only in the line wing, and 2) knowing the core is not modeled accurately, we do not include a large enough number of spatial eigenmodes in the sum to resolve it. However, an approximate fitting function dependent on parameters $a$ and $\tau_0$ is found that adequately represents the escape time distribution of the Monte Carlo results within these constraints.

Our characterization of the escape time distribution leads in to a possible application of this work. Models of the escape of hydrogen gas in exoplanet atmospheres can be constructed with a treatment of resonant scattering of \lya in spherical geometry. The Monte Carlo method can be used for this problem, but is limited by its high computational demand for large $\tau_0$ where there are many photon scatterings before escape. We seek to develop a computationally efficient random walk model which requires a solution in spherical geometry. There are several methods that are commonly used to accelerate Monte Carlo radiation transfer calculations, including core skipping methods \citep{1968ApJ...153..783A,2002ApJ...567..922A} and hybrid diffusion methods \citep{2018MNRAS.479.2065S}, but these methods are not sufficient for our application since they do not track the amount of time photons spend in the line core. Another approach with wide application is modified random walk methods, such as those discussed in \citet{1984JCoPh..54..508F, 2009A&A...497..155M, 2010A&A...520A..70R}; an outgoing photon is randomly sampled on the surface of the outgoing sphere by drawing its properties from distributions in outgoing frequencies, directions, and escape times, based on solutions to the diffusion equation. A method similar to this has been applied by \citet{2006ApJ...645..792T} to Lyman $\alpha$ transfer using the \citet{1990ApJ...350..216N} solution, but this solution of course does not utilize the frequency-dependent boundary condition at the surface of the sphere. Furthermore, to perform a full radiation hydrodynamic simulation with Monte Carlo acceleration, it will be necessary to calculate radiation forces due to \lya transfer on the faces of each grid cell in the simulation. Similar calculations have been done in \citet{1976ApJ...208..286W} in plane-parallel geometry. However, these solutions are limited to optical depths below $2.5 \times 10^3$. For this work, it would be necessary to model line center optical depths of up to 1 million or more. We will utilize the solution derived here as the basis for our own novel implementation of the modified random walk method.

@@

We have examined previous solutions to \lya transfer including resonant scattering in the limit of large optical depth, noting that the separation of variables and treatment of the boundary condition in \citet{1973MNRAS.162...43H}, \citet{1990ApJ...350..216N}, \citet{2006ApJ...649...14D} and others produces a discrepancy in the outgoing spectrum as compared with Monte Carlo. Here, we have derived the solution in spherical geometry with an appropriate treatment of the surface boundary condition. The key result is that the errors in the previously-cited works have been quantified via a correction term, $H_{\rm bc}$, which resolves an excess in flux at the spectral peak and a deficit in the line wing of the calculated spectrum of \lya radiation as compared with Monte Carlo. The size of $H_{\rm bc}/H_0$ is of order unity when the spectral peaks are near the Doppler core, and diminishes at larger $\tau_0$ following a $(a\tau_0)^{-1/3}$ scaling.

The time-dependent transfer equation is solved numerically with an eigenfunction expansion. We demonstrate that it agrees with the steady-state spectrum for $x_s=0$ when integrated over time, though its rate of numerical convergence is slow and requires a sum over many modes to become accurate. The time-dependent solution is utilized to create wait-time distributions for photons escaping the sphere of optically-thick hydrogen gas. We compare the calculations from the time-dependent solution with Monte Carlo for a sample of $\tau_0$, noting general agreement in the resulting escape time distributions. In future work, this will be used to accelerate Monte Carlo \lya transfer at large optical depths with potential applications in radiation hydrodynamic simulations of the atmospheres of exoplanets.

@@

This research was funded by NASA ATP grant 80NSSC18K0696, ``Exoplanetary MHD Outflows Driven by EUV Heating, Lyman alpha Radiation Forces and Stellar Tides". Resources supporting this work were provided by the NASA High-End Computing (HEC) Program through the NASA Advanced Supercomputing (NAS) Division at Ames Research Center.
\restartappendixnumbering

\software{\texttt{numpy} \citep{2020NumPy-Array}, \texttt{scipy} \citep{2020SciPy-NMeth}}
